---
title: "Plotting responses to STB isolates"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load tables of means
```{r}
dat <- read.csv("/Users/hafeeza/Documents/WatSeq/Analyses_for_paper/WatSeqStb_fulltable.csv", na.strings = "-", header = TRUE)
```
```{r, include=FALSE}
# remove unneeded columns
dat <- dat[,-c(5,9,10,14)]
```

```{r}
head(dat)
```

### Format data

Creating a column called 'SimpAlleles' which indicates whether lines have functional alleles of Stb6 or Stb15.
```{r}
dat[which(dat$Stb6=="1" | dat$Stb6=="2") , "SimpStb6"] <- "1"
dat[is.na(dat$SimpStb6), 'SimpStb6'] <- "0" 

dat[which(dat$Stb15=="A/R"), "SimpStb15"] <- "1"
dat[is.na(dat$SimpStb15), 'SimpStb15'] <- "0"

dat[which(dat$SimpStb6=="1"), 'SimpAlleles'] <- "Stb6"
dat[which(dat$SimpStb15=="1"), 'SimpAlleles'] <- "Stb15"
dat[which(dat$SimpStb15=="1" & dat$SimpStb6 =="1"), 'SimpAlleles'] <- "Stb15+Stb6"
dat[is.na(dat$SimpAlleles), 'SimpAlleles'] <- "None"
dat[1:21, 'Type'] <- "Cultivar"
dat$Line <- as.factor(dat$Line)

# Don't need Baj labelled in the figures so giving it a different category
dat[2,'Type'] <- 'Wheat'
```

Load packages needed
```{r}
if(!require(ggplot2)){install.packages(ggplot2)}
library(ggplot2)
if(!require(cowplot)){install.packages(cowplot)}
library(cowplot)
if(!require(ggpubr)){install.packages(ggpubr)}
library(ggpubr)
if(!require(ggrepel)){install.packages(ggrepel)}
library(ggrepel)
```

### Generate plot of pcynidia scores by allele

This plot is of logit pycnidia AUDPC scores of IPO323 and IPO88004 plotted against eachother, with the allele combination of each line indicated by its shape and colour.
```{r}
ggplot(dat, aes(y=LogitpAUDPC_IPO88004, x=LogitpAUDPC_IPO323, fill=SimpAlleles, shape=SimpAlleles)) +
  geom_point(aes(y=LogitpAUDPC_IPO88004, x=LogitpAUDPC_IPO323),size=2, colour="#5A5757") +
  scale_fill_manual(values=c("#ABADB3", "#FFB000", "#DC267F", "#785EF0")) +
  scale_shape_manual(values=c(25,22,23,21)) +
  xlab("Logit pAUDPC IPO323") + ylab("Logit pAUDPC IPO88004") +
  geom_label_repel(data = subset(dat, Type=="Cultivar"), aes(y=LogitpAUDPC_IPO88004, x=LogitpAUDPC_IPO323,label=Line, colour=SimpAlleles), min.segment.length=unit(0,'lines'), fill="white", show.legend = FALSE, nudge_y = 2, nudge_x = 3) +
  scale_colour_manual(values=c("#ABADB3", "#FFB000", "#DC267F", "#785EF0")) +
  theme_half_open() +
  background_grid()
  #ggsave("Pycplot_Stbgenes.png", height=5, width=6.5)
```
Reverse logit transformation of axes:
```{r}
al = c(-9, -8, -6, -4, -3, -2, -1, 0, 1, 2, 3, 4, 6, 8)

inv.logit <- function(y,a) {
  ((100+a)*exp(y) - a) / (1+exp(y))
}

zapsmall(inv.logit(al,a=0.5))
```

### Plot by place of origin

Load the origin/location data and remove missing values:
```{r}
org <- read.csv("/Users/hafeeza/Documents/WatSeq/Analyses_for_paper/watkins_origins.csv")
cs <- org[which(org$WatNo!="#N/A"), ]
head(cs)
```

Install ggmap, which uses Google Maps:
```{r}
if(!require(devtools)) install.packages("devtools",repos = "https://cran.ma.imperial.ac.uk/")
devtools::install_github("dkahle/ggmap", ref = "tidyup")
library(ggmap)
```
It is required to register a Google API key. This can be generated following instructions [here](https://support.google.com/googleapi/answer/6158862?hl=en).
```{r}
register_google(key = 
                "AIzaSyDKVWknyxoHsbdIkixqRLdUxnm7XUye7TQ")
```

Merge the location data with the rest of the data, removing cultivars:
```{r}
colnames(cs)[2] <- "Line"
dat2 <- merge(dat, cs, by="Line")
wonly <- dat2[-which(dat2$Type=="Cultivar"),]
```
Plot the landraces by longitude and latitude onto the map, with colours and shapes of points corresponding to their combination of Stb6 and Stb15 alleles:
```{r}
theme_set(theme_bw(16))
world_map <- map_data("world")
ggplot() + 
  geom_polygon(data=world_map, aes(x=long, y=lat,group=group), colour="#5A5757", fill="#EFEAED", linewidth=0.4) +
  geom_point(aes(x = Longitude, y = Latitude, fill = SimpAlleles, shape=SimpAlleles),
             colour="#5A5757", stroke=0.4, size=1.7, data = wonly) +
  scale_fill_manual(values=c("#ABADB3", "#FFB000", "#DC267F", "#785EF0")) +
   scale_shape_manual(values=c(25,22,23,21)) +
  theme_map(12)
#ggsave("map.png", width=24, height=13)
```

